Sequential rank agreement methods for comparison of ranked lists

### 2015-06-26 latex/paper.Rnw
Draft

To interpretate the sequential rank agreement scale we propose by
several different benchmark values. Our data free benchmark value is
obtained by computing the sequential rank agreement of $L$ lists each
containing a permutation of $P$ items. Practically this benchmark
value can be obtained by averaging the result of repeated computer
experiments. In the uncensored case each experiment contains $L$ lists
of random permutations of the numbers $1,\dots,P$. In the censored
case only the uncensored part of the orginal lists is randomly
permuted.

In our experience XX repetitions are sufficient to reduce Monte Carlo
error. Table Appendix A shows benchmark values obtained with

Our no-information benchmark can be applied in situations where each
list is an importance ranking of $P$ predictors resulting from
regression analyses. The idea is to repeat the regression analyses in
modified data where the outcome vector is randomly permuted. This
artificially removes any potential information regarding the
regression that the real data may contain.

- third approach?

\subsection{Independent lists}

\subsection{Permuted outcomes + analyses}

Do the following a large number of times
\begin{enumerate}
  \item Permute outcome vector
  \item Redo analyses for all $L$ methods
  \item Compute sequential rank agreement for
\end{enumerate}

Lidt spøjs ting man matcher dem op imod

@@

The overlap among $L$ lists observed until depth $d$ is defined as the
proportion of items that are found in all $L$ lists compared to the
possible number of items found in all lists observed until depth $d$:
\begin{equation}
O(d) = \frac{| \cap_{l=1}^L \{\pi_l(r); r\leq d \} |}{d}.
\end{equation}
Typically, the overlap is only considered for pairwise comparisons

The average overlap at depth $d$ is defined as the average of the
overlaps until $d$,
\begin{equation}
AO(d) = \sum_{i=1}^d \frac{O(d)}{d}.
\end{equation}
This definition directly ensures that extra emphasis is included on
the top of the list since the overlap of the items towards the top are
included in the majority of the calculations.

In this section we apply the method to the ColonData data from the R
package MAMA (Ihnatova, 2013), which consists of gene expression
measurements from a subset of 500 genes from 77 individuals from
Denmark, 36 from Australia and 41 from Japan as well as information on
microsattelite stability (stable MSS/unstable MSI) which is often used
as an indicator for colorectal cancer. The genes were tested for
difference in expression in the MSI/MSS groups using the limma package
(Smyth, 2005). The genes were then ranked according to p–values giving
three lists that we wish to compare, the top of these can be seen in
Table 4.

@@

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.45\textwidth]{paper-fig-ao1}
\includegraphics[width=.45\textwidth]{paper-fig-ao2}
\end{center}
 \caption{Average overlap (left figure) and sequential rank agreement
   (right figure) for two different analysis methods
   applied to the Golub data (thick red line). The red and blue shaded
 areas represent the pointwise 95\% normal areas for the XXX and XXY benchmarks,
respectively.}
 \label{fig:case1}
\end{figure}

@@

>>>>

To interpret the sequential rank agreement values we propose two
different benchmark values corresponding to two different
hypotheses. We wish to determine if we observe better agreement than
what would be expected if there were no relevant information available
in the data.

The first reference hypothesis is 
\begin{eqnarray*}
H_0  & : &  \text{The list rankings correspond to complete randomly}\\
       &  & \text{permuted lists} 
\end{eqnarray*}
which not only assumes that there is no information in the rankings
but also that the methods are completely independent.

Alternatively, we can remove the restriction on the independence among
the methods by only requiring that there is no information in the
ranking but that the rankings are all based on applying the 
method/approaches to the same data
\begin{eqnarray*}
H_1 & :&  \text{The list rankings are based on data containing}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}
$H_0$ is clearly the strictest (and also potentially quite
unrealistic) null hypothesis but we can easily obtain realizations
from that null hypothesis simply by permuting the items for each list
and then computing the sequential rank agreement for the permuted
lists. In the uncensored case each experiment contains $L$ lists of
random permutations of the numbers $1,\dots,P$. For the censored case
we permute the number as before but censor the same number of items
for each list as were censored in the original dataset. The sequential
rank agreement curve from the the original data can then be compared
to the pointwise rank agreements obtained from under $H_0$.

To obtain a reference distribution under $H_1$ we use a randomization
test approach on the original data to remove any associations to the
outcome in the data that was used to generate the rankings before
computing the ranking and subsequently the sequential rank agreement
from the randomized data. The idea is to repeat the ranking procedures
in the modified data where the outcome vector is randomly permuted. In
practice we permute the outcomes but keep any predictors before
ranking to preserve any structure in the predictors. This artificially
removes any potential information regarding the regression that the
real data may contain. The randomization approach requires that we
have the original data available, so we can compute the ranked lists
from the randomized data and not just the ranked lists and as such it
may not be possible to evaluate $H_1$ in all situations.

If the sequential rank agreement for the original data lies
substantially below the distribution of the sequential rank agreements
obtained under either $H_0$ or $H_1$ then this suggests that the
original ranked lists agree \emph{more} that we would expect if the
ranked lists were based on data with no information.

Figure~\ref{fig:example1} shows the sequential rank distributions
based on $400$ samples under $H_0$ and $H_1$, respectively. Not
surprisingly, the null distribution of $H_1$ has lower sequential rank
agreement than the null distribution of $H_0$ because the four methods
used to rank the data ($t$ test, logistic regression, elastic net, and
MIC) generally identify and rank the same set of predictors that show
spurious associations. The two bottom panels in
Figure~\ref{fig:example1} also indicate that the observed sequential
rank agreement is better than what would be expected for the full
dataset but that the censored data (the bottom-right plot) suggests
that there may be at most 1 or 2 ranked items that are towards the top
of the lists that yield a result better than what would be expected.

It is important to stress that neither $H_0$ nor $H_1$ are concerned
with whether the $L$ rankings are relevant. Both hypotheses are purely
considering how the rankings agree in a situation where there is no relevant
information available in the data used for creating the rankings.

@@

The average overlap is widely used in comparing (two) ranked list and
the idea behind the average overlap closely resembles the sequential
rank agreement \citep{Fagin2003,Webber2010}. Here we will present an
extention of the average overlap to more that two lists and compare this
to sequential rank agreement.

The overlap among $L$ lists observed until depth $d$ is defined as the
proportion of items that are found in all $L$ lists compared to the
possible number of items found in all lists observed until depth $d$:
\begin{equation}
O(d) = \frac{| \cap_{l=1}^L \{\pi_l(r); r\leq d \} |}{d}. \label{def:overlap}
\end{equation}
Typically, the overlap is only considered for pairwise comparisons but in
\eqref{def:overlap} we make no such distinction.

The average overlap at depth $d$ is defined as the average of the
overlaps until depth $d$,
\begin{equation}
AO(d) = \sum_{i=1}^d \frac{O(d)}{d}.
\end{equation}
This definition directly ensures that extra emphasis is included on
the top of the list since the overlap of the items towards the top are
included in the majority of the calculations.

There are three main drawbacks of the average overlap method: First it
is highly sensitive to items at the top of the list. The average
overlap profile changes dramatically depending on the occurrence of
similar items in the lists. Secondly, when the number of lists, $L$,
is high then it might be difficult to obtain a non-zero overlap (and
hence a non-zero average overlap) at the beginning of the lists because
the overlap does not increase until an item is present in all $L$
lists. Finally, the average overlap has an interpretation in terms of
moving averages of  percentages (of $L$ sets) which somewhat less
intuitive than simply the average rank distance that we propose.

In this section we compare the average overlap and the sequential rank
agreement to the Golub data  \citep{Golub1999}. To make it more
similar to the way the average ovelap is typically used we restrict
attention to only two analysis methods: $t$ tests and logistic
regression. Thus we have two marginal model approaches that are very
similar and the first two columns of Table~\ref{tab1} show the actual top-10
ranking of the predictors.

@@

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.45\textwidth]{paper-fig-ao1}
\includegraphics[width=.45\textwidth]{paper-fig-ao2}
\end{center}
 \caption{Average overlap (left figure) and sequential rank agreement
   (right figure) for comparing two different analysis methods
   (marginal $t$ test and logistic regression)
   applied to the Golub data. The black lines are based on the full
   set of predictors while the red lines have the two highest
   associated predictors removed
   from the data before computing the average overlap and sequential
   rank agreement.}
 \label{fig:case1}
\end{figure}

The black lines Figure~\ref{fig:case1} show the average overlap and
sequential rank agreement for these data. It is clear from both plots
that there is perfect agreement towards the top of the ranked lists:
the average overlap is 1 and the sequential rank agreement is zero.
However, if we remove those two items from the data (i.e.,
gene/predictor 2124 and 896) and redo the analyses then we get
substantially different curves for the average overlap but not for
sequential rank agreement (red lines in Figure~\ref{fig:case1}). For
the sequential rank agreement we get roughly the same estimate of
agreement from rank 3 as we did from the full dataset which is a nice
property. For the average overlap the curves completely change when
these two items are missing and give a visually different conclusion
about the ranking of the items from rank 3 downwards.

@@

Censor lists at significane level?
